Rheologic constraints on the upper mantle from five years of postseismic deformation following the El Mayor-Cucapah earthquake

### 2016-04-20 JGRSubmission/agujournaltemplate.tex
commit before push

\documentclass[draft,linenumbers]{AGUJournal}
\draftfalse
\usepackage{amssymb,amsmath,amsthm,amsfonts}
\journalname{JGR-Solid Earth}

%set up the header
%\pagestyle{fancy}
%\lhead{Trever Hines}
%\chead{El Mayor Postseismic}
%\rhead{\today}

%\setlength{\headheight}{15pt}
%\renewcommand\headrulewidth{1.0pt} % Size of the header rule

\begin{Document}
%% Title
%%------------------------------------------------------------------------------

\title{	
Rheologic constraints on the upper mantle from five years of postseismic deformation following the El Mayor-Cucapah earthquake}
\author{T. T. Hines\fnref{fn}}
\author{E. A. Hetland\fnref{fn}}
\fntext[fn]{Department of Earth and Environmental Sciences, University of Michigan, Ann Arbor, MI, USA}

\begin{abstract}
Five years of postseismic deformation following the  Mw7.2 El Mayor-Cucapah earthquake reveals transient deformation that decays back to its pre-earthquake trend after about 3 years at epicentral distances greater than $\sim200$ km. In the near-field, we observe rapid transience that decays to a sustained rate which exceeds its pre-earthquake trend.  We attempt to determine the mechanisms driving this observed deformation, where we consider afterslip at seismogenic depths and viscoelastic relaxation in the lower crust and upper mantle as candidate mechanisms.  We find that early, rapid, near-field deformation can be explained with afterslip on the portion of the fault that ruptured coseismically, while the later, sustained, near-field deformation can be explained with either continued afterslip in an effectively elastic lower crust, or lower crustal viscoelastic relaxation with a steady-state viscosity of $\sim10^{19}$ Pa s. The trend in far-field deformation is best explained with a transient viscosity of $\sim10^{18}$ Pa s in the upper mantle. We argue that a transient rheology in the mantle is preferable over a Maxwell rheology because it better predicts the decay of the postseismic deformation, and also because it does not conflict with the generally higher, steady-state viscosities inferred from studies of geophysical processes occurring over longer time-scales.
\end{abstract}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Ground deformation in the years following a large ($\gtrsim$Mw7) earthquake provides insight into the mechanical behaviour of the crust and upper mantle. It has long been recognized that interpretations of postseismic deformation can be ambiguous because multiple postseismic deformation mechanisms can have qualitatively similar surface expressions \citep[e.g.][]{Savage1990}. Thanks to the dense geodetic network deployed throughout the 2000s as part of the Plate Boundary Observatory, the April 4, 2010, Mw7.2 El Mayor-Cucapah earthquake in Baja California produced observable postseismic deformation at more GPS stations than any other earthquake in California to date. With such a large collection of data, we attempt to discern the mechanisms driving postseismic deformation following the El Mayor-Cucapah earthquake, where we consider both afterslip and viscoelastic relaxation in the lower crust and upper mantle as candidate mechanisms.

Previous studies which have modelled postseismic deformation following the El Mayor-Cucapah earthquake include \citet{Pollitz2012}, \citet{Gonzalez-ortega2014}, \citet{Spinler2015}, and \citet{Rollins2015}. Of these studies, \citet{Gonzalez-ortega2014} and \citet{Rollins2015} have attempted to describe the postseismic deformation with afterslip in an elastic half-space.  \citet{Gonzalez-ortega2014} described five months of postseismic deformation, observed by near-field ($\lesssim$ 50 km from the rupture) campaign GPS and InSAR, with afterslip and contraction on the coseismically ruptured fault. \citet{Gonzalez-ortega2014} noted that their preferred model underestimated the GPS displacements for stations $\gtrsim 25$ km from the rupture and suggested that it could be the result of unmodelled viscoelastic relaxation.  Using continuous GPS stations, which are mostly north of rupture zone, \citet{Rollins2015} found that three years of postseismic deformation can be adequately explained by afterslip, albeit with an implausibly large amount of slip inferred on the least constrained, southern-most fault segment. Here, we suggest the afterslip inferred by \citet{Rollins2015} may have been acting as a proxy for distributed relaxation in the upper mantle.

\citet{Pollitz2012}, \citet{Rollins2015} and \citet{Spinler2015} have explored viscoelastic relaxation in the lower crust and upper mantle as a potential postseismic deformation mechanism. The rheology of the crust and mantle is largely unknown and so modelling postseismic deformation with viscoelastic relaxation requires one to assume a rheologic model and then find the best fitting rheologic parameters.  The inference of these rheologic parameters is a computationally expensive non-linear inverse problem which is typically approached with a forward modelling grid search method.  Consequently, a simplified structure for the lithosphere must be made to minimize the number of rheologic parameters that need to be estimated.  For example, it is commonly assumed that the lithosphere consists of only three homogeneous, Maxwell viscoelastic layers, which may be an inadequate representation of the lithosphere \citep{Riva2009,Hines2013}. To further reduce the model space dimensions being search, it is also necessary to make simplifying assumptions about the nature of afterslip. For example, one can assume a frictional model for afterslip and parametrize afterslip in terms of the unknown rheologic properties of the fault \citep[e.g.][]{Johnson2009,Johnson2004}. One can also assume that afterslip does not persist for more than a few months and then model the later postseismic deformation assuming it to be the result of only viscoelastic relaxation \citep[e.g.][]{Pollitz2012,Spinler2015}. However, postseismic afterslip in similar tectonic settings has been observed to persist for decades following earthquakes \citep{Cakir2012,Cetin2014}. Neglecting to allow for sustained afterslip as a postseismic mechanism could then lead to a biased inference of lithospheric viscosity. Indeed, the preferred viscoelastic model from \citet{Pollitz2012} significantly underestimates near-field deformation, which could be indicative of unmodelled continued afterslip.

In this study, we assume that both afterslip and viscoelastic relaxation can contribute to postseismic deformation. Modelling both of these mechanisms creates a high dimensional model space that must be searched with non-linear optimization methods. We first develop an initial postseismic model using the method described in \citet{Hines2016}. This method simultaneously estimate the afterslip and effective lithospheric viscosity structure necessary to describe the first 0.8 years of postseismic deformation following the El Mayor-Cucapah earthquake.  We then use the inferred effective viscosity structure to create a suite of postseismic models which are tested against the available 5 years of postseismic data.  Of the suite of models tested, we find that postseismic deformation following the El Mayor-Cucapah earthquake can be explained with afterslip beneath the Sierra Cucapah, a Zener rheology upper mantle with a transient viscosity that decays from $5\times10^{18}$ Pa s to $1\times10^{18}$ Pa s at 120 km depth, and a relatively stronger lower crust.

\begin{figure}
\includegraphics[scale=0.7]{Figures/ContextMap} 
\caption{Map of the region considered in this study.  The large focal mechanism is for the El Mayor-Cucapah earthquake and the three small focal mechanisms are for the Ocotillo earthquake and the two main shocks during the Brawley swarm.  The black dots indicate the locations of GPS stations used in this study.  The fault geometry used in this study is shown in cyan where dashed lines indicate buried edges of the fault segments.  The green and red boxes demarcate the extent of the near-field and far-field maps (figures \ref{fig:NearField} and \ref{fig:FarField}).  Stations within the blue sector, which highlights the area within 10$^\circ$ of the El Mayor-Cucapah P axis, are used in the record sections in Figures \ref{fig:RecordSectionElastic}, \ref{fig:RecordSection1}, and \ref{fig:RecordSection2}}.       
\label{fig:ContextMap}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data Processing}\label{sec:Data}

We use continuous GPS position time series provided by University Navstar Consortium (UNAVCO) for stations within a 400 km radius about the El Mayor-Cucapah epicentre. We collectively describe the coseismic and postseismic displacements resulting from the El Mayor-Cucapah earthquake as $u_\mathrm{post}(t)$.  We consider the GPS position time series, $u_\mathrm{obs}(t)$, to be the superposition of $u_\mathrm{post}(t)$, secular tectonic deformation, annual and semi-annual oscillations, and coseismic offsets from significant earthquakes over the time span of this study.  The June 14, 2010, Mw5.8 Ocotillo earthquake and the Brawley swarm, which included an Mw5.5 and an Mw5.3 event on August 26, 2012, (Figure \ref{fig:ContextMap}) are the only earthquakes that produced noticeable displacements in any of the time series.  We treat the displacements resulting from the Brawley swarm as a single event because the time series are provided by UNAVCO as daily solutions. Although the Ocotillo earthquake had its own series of aftershocks \citep{Hauksson2011}, neither the Ocotillo earthquake nor the Brawley swarm produced detectable postseismic deformation and we model displacements resulting from these events with a Heaviside function, $H(t)$.  We then model $u_\mathrm{obs}(t)$ as 
\begin{equation}
  u_\mathrm{obs}(t) = u_\mathrm{pred}(t) + \epsilon
\end{equation}
where
\begin{equation}\label{TimeSeriesModel}
  \begin{split}  
    u_\mathrm{pred}(t) = &u_\mathrm{post}(t)H(t-t_\mathrm{emc}) + c_0 + c_1t + \\
                         &c_2\sin(2\pi t) + c_3\cos(2\pi t) + c_4\sin(4\pi t) + c_5\cos(4\pi t) + \\
                         &c_6H(t-t_\mathrm{oc}) + c_7H(t-t_\mathrm{bs}).
  \end{split}
\end{equation}
In the above equations, $t_\mathrm{emc}$, $t_\mathrm{oc}$ and $t_\mathrm{bs}$ are the times of the El Mayor-Cucapah earthquake, Ocotillo earthquake, and the Brawley swarm, respectively, $c_0$ through $c_7$ are unknown coefficients, and $\epsilon$ is the observation noise. We only estimate jumps associated with the Ocotillo earthquake and Brawley swarm for stations within 40 km of their epicentres.

Stations which recorded displacements that clearly cannot be described by the aforementioned processes are not included in our analysis. This includes stations in the Los Angeles basin, where anthropogenic deformation can be larger than the postseismic signal that we are trying to estimate \citep{Bawden2001,Argus2005} . In order to ensure an accurate estimation of the secular deformation, we only use stations that were installed at least six months prior to El Mayor-Cucapah earthquake. Several GPS stations were installed after the El Mayor-Cucapah earthquake to improve the spatial resolution of postseismic deformation \citep{Spinler2015} and it is possible to subtract secular velocities derived from elastic block models \citep[e.g.][]{Meade2005} from velocities recorded at the newly installed stations to get an estimate of postseismic velocities. However, we use coseismic and postseismic displacements, rather than velocities, in our inverse method described in Section \ref{sec:Model}. We use displacements because estimating velocities from an already noisy displacement time series can introduce significant aleatoric and epistemic uncertainties depending on exactly how the estimation is done. This modelling choice prevents us from using the newly installed stations in Baja California for our analysis.

The October 16, 1999, Mw7.1 Hector Mine earthquake, which occurred about 270 km north of the El Mayor-Cucapah epicentre, produced transient postseismic deformation which we do not wish to model, either mechanically or through empirical line fitting.  We thus restrict our analysis to deformation observed six years after the Hector Mine earthquake, which is when postseismic velocities at sites proximal to the Hector Mine epicentre are approximately constant \citep{Savage2009}. When appraising our model fit in Section \ref{sec:Model}, we see some systematic residuals in the vicinity of the Hector Mine epicentre, which may be the result of errors in the assumption that the trend in Hector Mine postseismic deformation is linear after six years.

@@

The smoothness of $u_\mathrm{post}(t)$ is controlled by the chosen value of $\sigma^2$, which describes how rapidly we expect postseismic displacements to vary over time.  Setting $\sigma^2$ equal to zero will effectively result in modelling $u_\mathrm{post}(t)$ as a straight line which is insufficient to describe the expected transient behaviour in postseismic deformation. The other end member, where $\sigma^2$ is infinitely large, will result in $u_\mathrm{pred}(t)$ over fitting the data. While one can use a maximum likelihood based approach for picking $\sigma^2$ \citep[e.g.][]{Segall1997}, we rather take a subjective approach and choose a value for $\sigma^2$ that is just large enough to faithfully describe the observed deformation at the most near-field station in our study, P496, which exhibits the most pronounced rapid changes in velocity. This ensures that $\sigma^2$ will be sufficiently large so that our estimate of $u_\mathrm{post}(t)$ does not smooth out potentially valuable postseismic signal at the remaining stations. We find that using $\sigma^2 = 0.05 \mathrm{m}^2 / \mathrm{yr}^3$ adequately describe all but the first week of postseismic deformation at station P496, which gets incorporated into our estimate of coseismic displacements (Figure \ref{fig:P496}). By down sampling, we implicitly assume that the first week of deformation is overwhelmingly the result of afterslip and this afterslip is lumped into our estimates of coseismic slip. \citet{Freed2007a} noted that postseismic deformation can be observed at distances greater than 200 km from the Hector Mine earthquake and, after filtering the time series for stations up to 400 km from the El Mayor-Cucapah epicentre, we also clearly see far reaching postseismic transient deformation (Figure \ref{fig:P619}).

\begin{figure}
\includegraphics[scale=0.7]{Figures/FilterP496}
\centering
\caption{Left panels show GPS time series from UNAVCO (black) and the predicted displacement (blue) from eq. (\ref{TimeSeriesModel}) for a near-field station.  Red lines indicate the times of the El Mayor-Cucapah and Ocotillo earthquake. The right panels show estimated coseismic and postseismic displacements, $u_\mathrm{post}$, which are extracted from the predicted displacements.  The 68\% confidence interval is shown in light blue.}
\label{fig:P496}
\end{figure}

\begin{figure}
\includegraphics[scale=0.7]{Figures/FilterP619}
\centering
\caption{same as Figure \ref{fig:P496} but for a far-field station.} 
\label{fig:P619}
\end{figure}

It is important to note that the shown uncertainties in $u_\mathrm{post}(t)$ do not account for the non-negligible epistemic uncertainty in eq. (\ref{TimeSeriesModel}).  For example, we assume a constant rate of secular deformation, which appears to be an appropriate approximation for all but perhaps the stations closest to the Hector Mine epicentre, as noted above.  Also, our model for seasonal deformation in eq. (\ref{TimeSeriesModel}) assumes a constant amplitude over time, which means that any yearly variability in the climatic conditions could introduce systematic residuals \citep{Davis2012}. Indeed, it would be more appropriate to consider the seasonal amplitudes $c_2-c_5$ in eq. (\ref{TimeSeriesModel}) as stochastic variables \citep{Murray2005}. By using constant seasonal amplitudes our estimate of $u_\mathrm{post}(t)$ seems to describe some of the unmodelled annual and semi-annual oscillations (e.g. Figure \ref{fig:P619}).

We show in Figures \ref{fig:NearField} and \ref{fig:FarField} the near and far-field postseismic displacements accumulated over the intervals  0-0.8 years, 0.8-3.0 years, and 3.0-5.0 years, as well as the coseismic displacements.  In the far-field to the north, we can see south trending displacements for the first 3 years following the earthquake at stations as far as $\sim$400 km from of the El Mayor Cucapah epicentre.  These displacements are most pronounced along the direction of the El Mayor-Cucapah P axis. A similar eastward trend can be seen in the few far-field stations along the T axis in Arizona.  After 3 years, the trend in far-field postseismic deformation is barely perceptible.  The vertical deformation in the far-field is difficult to attribute to postseismic processes.  Most far-field stations display an initial subsidence for the first year after the El Mayor-Cucapah earthquake followed by continued uplift.  This trend in vertical deformation can be observed in all three of the quadrants where postseismic data is available, which means that the vertical deformation does not exhibit an anti-symmetric quadrant pattern, as would be expected for postseismic processes.  Although we use vertical deformation in our analysis in Section \ref{sec:Model},  we do not put an emphasis on trying to describe the vertical deformation as it likely does not have postseismic origins.

\begin{figure}
\includegraphics[scale=0.7]{Figures/MapViewNearField}
\centering 
\caption{Near-field coseismic and postseismic displacements (black) as well as predicted displacements for our preferred model from Section \ref{sec:FullInversion} (green).  Observed vertical deformation is shown as an interpolant and predicted vertical displacements are shown within the circles at the base of each vector.  Note that the interpolant is not well constrained in Mexico where there is no data available.}
\label{fig:NearField}
\end{figure}

\begin{figure}
\includegraphics[scale=0.7]{Figures/MapViewFarField}
\centering 
\caption{Same as Figure \ref{fig:NearField} but for far-field stations.}
\label{fig:FarField}
\end{figure}

>>>>

%% 11/23/2015
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% AGUJournalTemplate.tex: this template file is for articles formatted with LaTeX
%
% This file includes commands and instructions
% given in the order necessary to produce a final output that will
% satisfy AGU requirements. 
%
% You may copy this file and give it your
% article name, and enter your text.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PLEASE DO NOT USE YOUR OWN MACROS
% DO NOT USE \newcommand, \renewcommand, or \def, etc.
%
% FOR FIGURES, DO NOT USE \psfrag or \subfigure.
% DO NOT USE \psfrag or \subfigure commands.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% All questions should be e-mailed to latex@agu.org.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Step 1: Set the \documentclass
%
% There are two options for article format:
%
% 1) PLEASE USE THE DRAFT OPTION TO SUBMIT YOUR PAPERS.
% The draft option produces double spaced output.
% 
% 2) numberline will give you line numbers.

%% To submit your paper:
\documentclass[draft,linenumbers]{AGUJournal}
\draftfalse
\usepackage{amssymb,amsmath,amsthm,amsfonts}
%% For final version.
% \documentclass{AGUJournal}

\journalname{JGR-Solid Earth}

\begin{document}

\title{Rheologic constraints on the upper mantle from five years of postseismic deformation following the El Mayor-Cucapah earthquake}

\authors{T. T. Hines\affil{1}, E. A. Hetland\affil{1}}

\affiliation{1}{Department of Earth and Environmental Sciences, University of Michigan, Ann Arbor, Michigan, USA.}

\correspondingauthor{T. T. Hines}{hinest@umich.edu}

%% Keypoints, final entry on title page.

% Example: 
% \begin{keypoints}
% \item	List up to three key points (at least one is required)
% \item	Key Points summarize the main points and conclusions of the article
% \item	Each must be 100 characters or less with no special characters or punctuation 
% \end{keypoints}

%  List up to three key points (at least one is required)
%  Key Points summarize the main points and conclusions of the article
%  Each must be 100 characters or less with no special characters or punctuation

\begin{keypoints}
\item = enter point 1 here = 
\item = enter point 2 here = 
\item = enter point 3 here = 
\end{keypoints}

\begin{abstract}
Five years of postseismic deformation following the  Mw7.2 El Mayor-Cucapah earthquake reveals transient deformation that decays back to its pre-earthquake trend after ${\sim}3$ years at epicentral distances greater than ${\sim}200$ km. At closer distances, the rapid transience decays to a sustained rate which exceeds its pre-earthquake trend.  We attempt to determine the mechanisms driving this deformation, where we consider afterslip at seismogenic depths and viscoelastic relaxation in the lower crust and upper mantle as candidate mechanisms.  We find that early, rapid, near-field deformation can be explained with afterslip on the fault that ruptured coseismically, while the later, sustained, near-field deformation can be explained with either continued afterslip in an effectively elastic lower crust, or lower crustal viscoelastic relaxation with a steady-state viscosity of ${\sim}10^{19}$ Pa s. The trend in far-field deformation is best explained with a transient viscosity of ${\sim}10^{18}$ Pa s in the upper mantle. We argue that a transient rheology in the mantle is preferable over a Maxwell rheology because it better predicts the decay in postseismic deformation, and also because it does not conflict with the generally higher, steady-state viscosities inferred from studies of geophysical processes occurring over longer time-scales.
\end{abstract}

\section{Introduction}
Ground deformation in the years following a large ($\gtrsim$Mw7) earthquake provides insight into the mechanical behaviour of the crust and upper mantle. It has long been recognized that interpretations of postseismic deformation can be ambiguous because multiple postseismic deformation mechanisms can have qualitatively similar surface expressions \citep[e.g.][]{Savage1990}. Owing to the dense geodetic network deployed throughout the 2000s as part of the Plate Boundary Observatory, the postseismic deformation following the April 4, 2010, Mw7.2 El Mayor-Cucapah earthquake in Baja California was observed at more GPS stations than any other earthquake in California to date. With such a large collection of data, we attempt to discern the mechanisms driving postseismic deformation following the El Mayor-Cucapah earthquake, where we consider both afterslip and viscoelastic relaxation in the lower crust and upper mantle as candidate mechanisms.

Previous studies which have modelled postseismic deformation following the El Mayor-Cucapah earthquake include \citet{Pollitz2012}, \citet{Gonzalez-ortega2014}, \citet{Spinler2015}, and \citet{Rollins2015}. Of these studies, \citet{Gonzalez-ortega2014} and \citet{Rollins2015} have attempted to describe the postseismic deformation with afterslip in an elastic half-space.  \citet{Gonzalez-ortega2014} described five months of postseismic deformation, observed by InSAR and GPS within ${\sim}50$ km rupture, with afterslip and contraction on the coseismically ruptured fault. \citet{Gonzalez-ortega2014} noted that their preferred model underestimated the GPS displacements for stations $\gtrsim 25$ km from the rupture and suggested that it could be the result of unmodelled viscoelastic relaxation.  Using only continuous GPS stations, which are mostly north of the rupture zone, \citet{Rollins2015} found that three years of postseismic deformation can be adequately explained by afterslip, albeit with an implausibly large amount of slip inferred on the least constrained, southern-most fault segment. Here, we suggest the afterslip inferred by \citet{Rollins2015} may have been acting as a proxy for distributed relaxation in the upper mantle.

\citet{Pollitz2012}, \citet{Rollins2015} and \citet{Spinler2015} have explored viscoelastic relaxation in the lower crust and upper mantle as a potential postseismic deformation mechanism. The rheology of the crust and mantle is largely unknown and so modelling postseismic deformation with viscoelastic relaxation requires one to assume a rheologic model and then find the best fitting rheologic parameters.  The inference of these rheologic parameters is a computationally expensive non-linear inverse problem which is typically approached with a forward modelling grid search method.  Consequently, a simplified structure for the lithosphere must be assumed in order to minimize the number of rheologic parameters that need to be estimated.  For example, it is commonly assumed that the lithosphere consists of only three homogeneous, Maxwell viscoelastic layers, which may be an inadequate representation of the lithosphere \citep{Riva2009,Hines2013}. To further reduce the dimensions of the model space, it is also necessary to make simplifying assumptions about the nature of afterslip. For example, one can assume a frictional model for afterslip and parametrize afterslip in terms of the unknown rheologic properties of the fault \citep[e.g.][]{Johnson2009,Johnson2004}. One can also assume that afterslip does not persist for more than a few months and then model the later postseismic deformation assuming it to be the result of only viscoelastic relaxation \citep[e.g.][]{Pollitz2012,Spinler2015}. However, postseismic afterslip in similar tectonic settings has been observed to persist for decades following earthquakes \citep{Cakir2012,Cetin2014}. Indeed, the preferred viscoelastic model from \citet{Pollitz2012} significantly underestimates deformation in the Imperial Valley, which could be indicative of unmodelled continued afterslip.  Neglecting to allow for sustained afterslip as a postseismic mechanism could then lead to biased inferences of lithospheric viscosities.

In this study, we assume that both afterslip and viscoelastic relaxation can contribute to postseismic deformation. Modelling both of these mechanisms creates a high dimensional model space that must be searched with non-linear optimization methods. We first develop an initial postseismic model using the method described in \citet{Hines2016}. This method simultaneously estimates the afterslip and effective lithospheric viscosity structure necessary to describe early postseismic deformation.  Our initial model describes the first 0.8 years of postseismic deformation following the El Mayor-Cucapah earthquake.  We then use the inferred effective viscosity structure from the initial model to create a suite of postseismic models to explain the 5 years of postseismic data available to date.  Of the suite of models tested, we find that postseismic deformation following the El Mayor-Cucapah earthquake can be explained with afterslip on a fault segment running through the Sierra Cucapah, a Zener rheology in the upper mantle with a transient viscosity that decays from $5\times10^{18}$ Pa s to $1\times10^{18}$ Pa s at 120 km depth, and a relatively stronger lower crust.

\begin{figure}
\includegraphics[scale=0.8]{Figures/ContextMap} 
\caption{Map of the region considered in this study.  The large focal mechanism is for the El Mayor-Cucapah earthquake and the three small focal mechanisms are for the Ocotillo earthquake and the two main shocks during the Brawley swarm.  The black dots indicate the locations of GPS stations used in this study.  The fault geometry used in this study is shown in cyan where dashed lines indicate buried edges of the fault segments.  The green and red boxes demarcate the extent of the near-field and far-field maps (figures \ref{fig:NearField} and \ref{fig:FarField}).  Stations within the blue sector, which highlights the area within 10$^\circ$ of the El Mayor-Cucapah P axis, are used in the record sections in Figures \ref{fig:RecordSectionElastic}, \ref{fig:RecordSection1}, and \ref{fig:RecordSection2}}.       
\label{fig:ContextMap}
\end{figure}

\section{Data Processing}\label{sec:Data}

We use continuous GPS position time series provided by University Navstar Consortium (UNAVCO) for stations within a 400 km radius about the El Mayor-Cucapah epicentre. We collectively describe the coseismic and postseismic displacements resulting from the El Mayor-Cucapah earthquake as $u_\mathrm{post}(t)$.  We consider the GPS position time series, $u_\mathrm{obs}(t)$, to be the superposition of $u_\mathrm{post}(t)$, secular tectonic deformation, annual and semi-annual oscillations, and coseismic offsets from significant earthquakes over the time span of this study.  The June 14, 2010, Mw5.8 Ocotillo earthquake and the Brawley swarm, which included an Mw5.5 and an Mw5.3 event on August 26, 2012, (Figure \ref{fig:ContextMap}) are the only earthquakes that produced noticeable displacements in any of the time series.  We treat the displacements resulting from the Brawley swarm as a single event because the daily solutions provided by UNAVCO cannot resolve the separate events.  Although the Ocotillo earthquake had its own series of aftershocks \citep{Hauksson2011}, neither the Ocotillo earthquake nor the Brawley swarm produced detectable postseismic deformation, and we model displacements resulting from these events with only a Heaviside function, $H(t)$, describing the coseismic offsets.  We then model $u_\mathrm{obs}(t)$ as 
\begin{equation}
  u_\mathrm{obs}(t) = u_\mathrm{pred}(t) + \epsilon,
\end{equation}
where
\begin{equation}\label{TimeSeriesModel}
  \begin{split}  
    u_\mathrm{pred}(t) = &u_\mathrm{post}(t)H(t-t_\mathrm{emc}) + c_0 + c_1t + \\
                         &c_2\sin(2\pi t) + c_3\cos(2\pi t) + c_4\sin(4\pi t) + c_5\cos(4\pi t) + \\
                         &c_6H(t-t_\mathrm{oc}) + c_7H(t-t_\mathrm{bs}).
  \end{split}
\end{equation}
In the above equations, $t_\mathrm{emc}$, $t_\mathrm{oc}$ and $t_\mathrm{bs}$ are the times of the El Mayor-Cucapah earthquake, Ocotillo earthquake, and the Brawley swarm, respectively, $c_0$ through $c_7$ are unknown coefficients, and $\epsilon$ is the observation noise. We only estimate jumps associated with the Ocotillo earthquake and Brawley swarm for stations within 40 km of their epicentres.

Stations which recorded displacements that clearly cannot be described by the aforementioned processes are not included in our analysis. This includes stations in the Los Angeles basin, where anthropogenic deformation can be larger than the postseismic signal that we are trying to estimate \citep{Bawden2001,Argus2005}. In order to ensure an accurate estimation of the secular deformation, we only use stations that were installed at least six months prior to El Mayor-Cucapah earthquake. Several GPS stations were installed after the El Mayor-Cucapah earthquake to improve the spatial resolution of postseismic deformation \citep{Spinler2015}.  Although it would be possible to subtract secular velocities derived from elastic block models \citep[e.g.][]{Meade2005} from velocities recorded at the newly installed stations to get an estimate of postseismic velocities, we do not do so here.  We use coseismic and postseismic displacements, rather than velocities, in our inverse method described in Section \ref{sec:Model}. We use displacements because estimating velocities from an already noisy displacement time series can introduce significant uncertainties depending on exactly how the estimation is done.  This choice prevents us from using the newly installed stations in Baja California for our analysis.

The October 16, 1999, Mw7.1 Hector Mine earthquake, which occurred ${\sim}270$ km north of the El Mayor-Cucapah epicentre, produced transient postseismic deformation which we do not wish to model, either mechanically or through empirical line fitting.  We thus restrict our analysis to deformation observed six years after the Hector Mine earthquake, which is when postseismic velocities at sites proximal to the Hector Mine epicentre are approximately constant \citep{Savage2009}. When appraising our model fit in Section \ref{sec:Model}, we see some systematic residuals in the vicinity of the Hector Mine epicentre, which may be the result of errors in the assumption that the trend in Hector Mine postseismic deformation is linear after six years.

@@

The smoothness of $u_\mathrm{post}(t)$ is controlled by the chosen value of $\sigma^2$, which describes how rapidly we expect postseismic displacements to vary over time.  Setting $\sigma^2$ equal to zero will effectively result in modelling $u_\mathrm{post}(t)$ as a straight line which is insufficient to describe the expected transient behaviour in postseismic deformation. The other end member, where $\sigma^2$ is infinitely large, will result in $u_\mathrm{pred}(t)$ over fitting the data. While one can use a maximum likelihood based approach for picking $\sigma^2$ \citep[e.g.][]{Segall1997}, we instead take a subjective approach and choose a value for $\sigma^2$ that is just large enough to faithfully describe the observed deformation at the most near-field station in our study, P496, which exhibits the most pronounced rapid changes in velocity. This ensures that $\sigma^2$ will be sufficiently large so that our estimate of $u_\mathrm{post}(t)$ does not smooth out potentially valuable postseismic signal at the remaining stations. We find that using $\sigma^2 = 0.05 \mathrm{m}^2 / \mathrm{yr}^3$ adequately describe all but the first week of postseismic deformation at station P496, which gets incorporated into our estimate of coseismic displacements (Figure \ref{fig:P496}). By down sampling, we implicitly assume that the first week of deformation is overwhelmingly the result of afterslip and this afterslip is lumped into our estimates of coseismic slip. \citet{Freed2007a} noted that postseismic deformation can be observed at distances greater than 200 km from the Hector Mine earthquake and, after filtering the time series for stations up to 400 km from the El Mayor-Cucapah epicentre, we also clearly see far reaching postseismic transient deformation (Figure \ref{fig:P619}).

\begin{figure}
\includegraphics[scale=0.8]{Figures/FilterP496}
\centering
\caption{Left panels show GPS time series from UNAVCO (black) and the predicted displacement (blue) from eq. (\ref{TimeSeriesModel}) for a near-field station.  Red lines indicate the times of the El Mayor-Cucapah and Ocotillo earthquake. The right panels show estimated coseismic and postseismic displacements, $u_\mathrm{post}$, which are extracted from the predicted displacements.  The 68\% confidence interval is shown in light blue.}
\label{fig:P496}
\end{figure}

\begin{figure}
\includegraphics[scale=0.8]{Figures/FilterP619}
\centering
\caption{same as Figure \ref{fig:P496} but for a far-field station.} 
\label{fig:P619}
\end{figure}

It is important to note that the shown uncertainties in $u_\mathrm{post}(t)$ do not account for the non-negligible epistemic uncertainty in eq. (\ref{TimeSeriesModel}).  For example, we assume a constant rate of secular deformation, which appears to be an appropriate approximation for all but perhaps the stations closest to the Hector Mine epicentre, as noted above.  Also, our model for seasonal deformation in eq. (\ref{TimeSeriesModel}) assumes a constant amplitude over time, which means that any yearly variability in the climatic conditions could introduce systematic residuals \citep{Davis2012}. Indeed, it would be more appropriate to consider the seasonal amplitudes $c_2-c_5$ in eq. (\ref{TimeSeriesModel}) as stochastic variables \citep{Murray2005}. By using constant seasonal amplitudes, our estimate of $u_\mathrm{post}(t)$ seems to describe some of the unmodelled annual and semi-annual oscillations (e.g. Figure \ref{fig:P619}).

We show in Figures \ref{fig:NearField} and \ref{fig:FarField} the near and far-field postseismic displacements accumulated over the time intervals 0-1 years, 1-3 years, and 3-5 years, as well as the coseismic displacements. 
Stations at epicentral distances ranging from ${\sim}200$ km to as far as ${\sim}400$ km have an elevated rate of deformation for the first 3 years following the earthquake.  This far-field deformation has a southward trend along the direction of the El Mayor-Cucapah P axis and a similar eastward trend can be seen in the few far-field stations in Arizona, located along the T axis.  After 3 years, the trend in far-field postseismic deformation is barely perceptible.  The vertical deformation in the far-field is difficult to attribute to postseismic processes.  Most far-field stations display an initial subsidence for the first year after the El Mayor-Cucapah earthquake followed by continued uplift.  This trend in vertical deformation can be observed in all three of the quadrants where postseismic data is available, which means that the vertical deformation does not exhibit an anti-symmetric quadrant pattern, as would be expected for postseismic processes.  Although we use vertical deformation in our analysis in Section \ref{sec:Model},  we do not put an emphasis on trying to describe the vertical deformation as it likely does not have postseismic origins.

\begin{figure}
\includegraphics[scale=0.8]{Figures/MapViewNearField}
\centering 
\caption{Near-field coseismic and cumulative postseismic displacements over the indicated time periods (black) as well as predicted displacements for our preferred model from Section \ref{sec:FullInversion} (green).  Observed vertical deformation is shown as an interpolated field and predicted vertical displacements are shown within the circles at the base of each vector.  Note that the interpolant is not well constrained in Mexico where there is no data available.}
\label{fig:NearField}
\end{figure}

\begin{figure}
\includegraphics[scale=0.8]{Figures/MapViewFarField}
\centering 
\caption{Same as Figure \ref{fig:NearField} but for far-field stations.}
\label{fig:FarField}
\end{figure}

The near-field postseismic deformation is notably sustained when compared to the far-field deformation.  Namely, the station in this study which is closest to the El Mayor-Cucapah epicentre, P496, has been moving at a steady rate of ${\sim}1.5$ cm/yr to the south since ${\sim}1$ year after the El Mayor-Cucapah earthquake.  Vertical postseismic deformation in the near-field does display a quadrant pattern which is consistent with the coseismic vertical deformation, suggesting that it is resulting from tectonic processes.  However, the vertical postseismic signal is only apparent for the first year after the earthquake (Figure \ref{fig:NearField}).  As with the far-field deformation, there is a general trend of uplift in the near-field after ${\sim}1$ year, which we do not consider to be related to postseismic processes.

\section{Postseismic Modeling}\label{sec:Model}
